---
title: "Weekly 2: Coffee Ratings"
author: "Kay Hales"
format:
  html:
    embed-resources: true
editor: visual
---

## Overview

You want to know which producers grow the best-rated coffee.

That sounds like a `group_by()` and a `summarise()`. The problem is that `producer` was typed by different people in different ways, so the same producer can appear under several labels, and some labels are not producers at all.

Your job is to work out how much of that you can fix, how much you shouldn't, and what that does to the answer.

A few things:

1.  Add `#| echo: false` to your code blocks.
2.  Use a new block when you move to a new task.
3.  Code goes in the blocks. **Answers go outside them**, as prose. A comment inside a code chunk is not an answer.
4.  Keep the original `producer` column. Every cleaned version goes in a **new** column.

Due Monday at 8pm.

## Your data

Import `coffee_ratings.csv` and inspect it.

```{r}
#| warning: false
library(tidyverse)

coffee <- read_csv("data/week3/weekly3/coffee_ratings.csv")

glimpse(coffee)
```

## Before you start

> What does one row represent?

One graded coffee sample.

> Is there a column that uniquely identifies a row? How would you check?

Not this time. `nrow(coffee)` and `nrow(distinct(coffee))` are both 1,139.

> Is `producer` a key? What does it mean that the same producer can appear on many rows?

No. One producer appears on many rows, so producer-to-row is one-to-many.

## How many producers?

Count the distinct values of `producer`.

```{r}
#| warning: false
coffee %>%
  count(producer, sort = TRUE)

n_distinct(coffee$producer)
sum(is.na(coffee$producer))

```

> How many distinct values are there?

692. 

> How many rows have no producer recorded? Does your distinct count include "missing" as one of its values? How would you check?

231. 

> Why shouldn't you read that number as the number of distinct producers?

It's counting NA values, but look at the output of your count. `La Plata` and `LA PLATA`.

## Normalize the representation

Create a new column, `producer_clean`, that makes the producer labels consistent in case and whitespace. Keep `producer`.

```{r}
#| warning: false
coffee_clean <- coffee |> 
  mutate(
    producer_clean = producer |> 
      str_to_upper() |> 
      str_squish()
  )

n_distinct(coffee$producer)
n_distinct(coffee_clean$producer)
```

> How many distinct values does `producer_clean` have?

666. 

> Is that the number of producers now? What does the reduction establish, and what does it *not* establish?

No. There are still NA values. And we can't guarantee that this has fixed all issues.

## Which labels actually merged?

A single number doesn't tell you what happened. Build a table showing every `producer_clean` value that more than one raw `producer` value was combined into, along with the raw values that went into it.

```{r}
#| warning: false
coffee_clean_prep <- coffee_clean |> 
  filter(!is.na(producer)) |> 
  distinct(producer, producer_clean) |> 
  count(producer_clean) |> 
  filter(n > 1) 

coffee_clean |> 
  distinct(producer, producer_clean) |> 
  semi_join(coffee_clean_prep, join_by(producer_clean)) |> 
  arrange(producer_clean, producer) |> 
  print(n = Inf)

# this let's us see the specifics of what labels became what exactly 
```

> How many cleaned labels absorbed more than one raw label?

22. 

> Look through them. Is every merge one you would stand behind?

For example, should we merge "VARIOUS" and "VARIOS"?

> Is this table more useful than knowing the number of labels went down? Why?

## Does it change anything?

Compare the La Plata coffees before and after normalizing. For each version, report the number of coffees and the mean and median `total_cup_points`.

```{r}
#| warning: false
coffee_clean |> 
  filter(producer_clean == "LA PLATA") |> 
  group_by(producer_clean) |> 
  summarise(
    n = n(), 
    mean_points = mean(total_cup_points, na.rm = TRUE),
    median_points = median(total_cup_points, na.rm = TRUE)
  )

coffee |> 
  filter(producer %in% c("La Plata", "LA PLATA")) |> 
  group_by(producer) |> 
  summarise(
    n = n(), 
    mean_points = mean(total_cup_points, na.rm = TRUE),
    median_points = median(total_cup_points, na.rm = TRUE)
  )
```

> What changed more: the number of coffees or the average score?

The number changed the most. The mean and median are fairly similar for this example.

> Suppose you had reported how many coffees La Plata produced using the raw column. How far off would you have been?

I would have been off by 30, or 9, depending on which one I selected.

## What case and whitespace didn't catch

Some labels still differ only in punctuation, i.e., a period, a comma, a slash.

Sort the distinct values of `producer_clean` and read through them. Near-duplicates end up next to each other.

```{r}
#| warning: false
coffee_clean |> 
  filter(!is.na(producer_clean)) |> 
  distinct(producer_clean) |> 
  arrange(producer_clean) |> 
  print(n = Inf)

```

> Find at least **three** pairs of labels that differ only in punctuation.

Now use exact replacement to make each pair consistent. Change only the values you identified and nothing else.

```{r}
#| warning: false

# our potential options
coffee_clean <- coffee_clean |>
  mutate(
    producer_exact = case_when(
      producer_clean == "AGRICOLA EL PILAR, SA." ~ "AGRICOLA EL PILAR, S.A.",
      producer_clean == "AGRICOLA NUEVA GRANADA, S.A" ~ "AGRICOLA NUEVA GRANADA, S.A.",
      producer_clean %in% c("CONQUISTA / MORITO", "CONQUISTA/ MORITO",
                            "CONQUISTAMORITO") ~ "CONQUISTA/MORITO",
      producer_clean == "KERCHANSHE TRADING P.L.C" ~ "KERCHANSHE TRADING PLC",
      producer_clean == "PLANTACIONES AGROPECUARIAS S.A" ~ "PLANTACIONES AGROPECUARIAS, S.A.",
      producer_clean == "SIEMBRAS VISION, S.A" ~ "SIEMBRAS VISION, S.A.",
      producer_clean == "TEÓFILO NARVÁEZ :" ~ "TEÓFILO NARVÁEZ",
      producer_clean == "YEH, LI CHUAN." ~ "YEH, LI CHUAN",
      TRUE ~ producer_clean
    )
  )
```

> For each change, say why you're confident it's the same producer.

> How long did it take to find those by reading? How many labels did you have to look at?

## The top producers

Using `producer_clean`, find the ten producers with the most coffees in the data.

```{r}
#| warning: false
coffee_clean |> 
  count(producer_clean, sort = TRUE) |> 
  slice_head(n = 10)
```

Look at that list carefully.

> Is every entry in it a producer? If not, which ones aren't, and what do they mean?

> Is any single producer split across more than one row of the list?

Now look at these four raw labels:

- Ipanema Agricola
- Ipanema Agricola S.A
- Ipanema Agrícola SA
- Ipanema Coffees

> Should these be combined into one producer? **Don't recode them.** Explain whether combining them would be normalization or classification, and what evidence you'd need before doing it.

> If you did combine them, what would happen to your top-ten list?

## What this means for your question

> A classmate says: "I reduced the producer column from 692 labels to 666, so the cleaned column is better." Is that justified? Explain.

> You set out to find which producers grow the best-rated coffee. Given everything above, what would you need to tell someone before they trusted your answer?

## Disclosure appendix

> What did you try first? Include the code you wrote and the actual error message, or a clear description of the problem if there was no error.

> If you used an AI system: the exact prompt, exactly what it returned, what it got wrong or left unclear, and what you changed.

> If you didn't use one, say so.
